---
title: "Data Import"
output: html_document
---
```{r}
library(tidyverse)
library(readxl)
library(haven)
```

Import our first dataset 
```{r}
litters_df =
  read_csv(file= "data_import_examples/FAS_litters.csv")

litters_df = janitor:: clean_names(litters_df)
```
Import your second dataset 
```{r}
pups_df =
  read_csv( "data_import_examples/FAS_pups.csv", skip = 3)
pups_df = janitor:: clean_names(pups_df)
```


### Look at the data 

```{r}
litters_df
pups_df
```

Head or tail 
```{r}
head(litters_df, 20)
tail(pups_df, 5)
```
```{r}
skimr::skim(pups_df)
```
```{r}
litters_df = 
    read_csv(
        file = "data_import_examples/FAS_litters.csv",
        na = c(".", "NA", ".")
        )
litters_df = janitor:: clean_names(litters_df)
```

```{r}
pups_df =
  read_csv(
    file ="data_import_examples/FAS_pups.csv", skip = 3,
    na = c(".", "NA", ".")
        )
pups_df = janitor:: clean_names(pups_df)
```

## Oh excel...

Jenny Bryan made "readxl" to solve our problems 
```{r}
mlb_df = read_excel("data_import_examples/mlb11.xlsx")
```
Look at the data 
```{r}
mlb_df
```
Load some Lotr data 

import Fellowship of the Ring words 
```{r}
fotr_df =
  read_excel(
    "data_import_examples/LotR_Words.xlsx",
    range = "B3:D6")
fotr_df
```
```{r}
tt_df =
  read_excel(
    "data_import_examples/LotR_Words.xlsx",
    range = "F3:H6")
tt_df
```

#Import SAS
Read in the PULSE dataset 

```{r}
pulse_df =
  read_sas("data_import_examples/public_pulse_data.sas7bdat")

pulse_df = janitor::clean_names(pulse_df)
```

#Whats so bad about "read.csv()
Import pups with "readcsv()
```{r}
pups_df_base =
  read.csv("data_import_examples/FAS_pups.csv")
  pups_df_base
```
Wont show: lazy about variable names: super slow for big dataset. 

```{r}
library(tidyverse)
```

```{r}
litters_df =
  read.csv(
    file= "data_import_examples/FAS_litters.csv",
    na = c(".", "NA", ".")
  )
litters_df = janitor::clean_names(litters_df)
pups_df =
  read_csv(
    file= "data_import_examples/FAS_pups.csv",
    skip =3,
    na = c(".", "NA", ".")
  )
pups_df = janitor::clean_names(pups_df)
```

## select some variables 

```{r}
select(litters_df, group, litter_number)
select(litters_df, group:gd_of_birth)
select(litters_df, -group)
select(litters_df, starts_with("gd"))
select(litters_df, starts_with("gd"), group)
select(litters_df, GROUP = group, everything())
select(litters_df, -starts_with("gd"))
```
a couple of alternaqtives 
```{r}
rename(litters_df, GROUP =group)
relocate(litters_df, litter_number, gd0_weight)
```

In the pups data, select litter_number, sex and PD ears 
```{r}
select(pups_df, litter_number, sex, pd_ears)
```
filer some rows 
```{r}
filter(litters_df, gd_of_birth == 19)
filter(litters_df, pups_born_alive <= 5)
filter(litters_df, pups_born_alive != 6)
filter(litters_df, group %in% c("Con7", "Con8"))
filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive <7)

filter(litters_df, pups_born_alive <4 | pups_born_alive >8)
drop_na(litters_df, gd0_weight)
```

In the pups dataset, keep only rows where PD walk is <11 and sex is 2 
```{r}
fikter(pups_df, pd_walk <11, sex == 2)
```
## mutate to change some columns 
lets add and change columns 

```{r}
mutate(
  litters_df,
  wt_gain =gd18_weight -gd0_weight, 
  wt_gain_squared =wt_gain ^2, 
  group+ str_to_lower(group)
)
```

```{r}
mutate(pups_df, pd_walk_minus_7 - pd_walk -7)
```

arrabnge data 
```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, desc(pups_born_alive))

```






